---
title: "Antarctica to Rockies Comparison"
output:
  html_document:
    toc: true
    toc_float: true
params:
  # TRUE  = re-run every lake's 01_ model script (and its 00_ data prep) before
  #         building the report, so changes to R/functions/ are picked up.
  # FALSE = skip the model runs and use the CSVs already in Data/model_outputs/.
  rerun_models: true
---

```{r setup, include = FALSE}
# Repository root. Every path below is built from it with file.path(), so the
# chunks work both when knitting and when run one at a time in RStudio
# (which runs chunks from this Rmd's folder, R/, not the repo root).
path <- path.expand("~/Library/CloudStorage/OneDrive-UCB-O365 (9-15-26 4:10 PM)/Documents/R-Repositories/TVLakes_IceModel")
if (!dir.exists(path)) stop("Repository folder not found: ", path)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 7)
```

```{r libraries}
source(file.path(path, "R/functions/libraries.R"))
source(file.path(path, "R/functions/functions.R"))
```

```{r lake-list}
# One row per lake: the model script that (re)creates its output, and the
# saved output this report reads. Add/remove lakes here.
lakes <- tibble::tribble(
  ~lake_key, ~model_script,                             ~output_file,
  "WLB",     "R/WLB/01_WLB_ice_thickness_model.R",      "Data/model_outputs/WLB_2017_2024_output.csv",
  "ELB",     "R/ELB/01_ELB_ice_thickness_model.R",      "Data/model_outputs/ELB_2017_2024_output.csv",
  "LH",      "R/LH/01_LH_ice_thickness_model.R",        "Data/model_outputs/LH_2017_2024_output.csv",
  "LF",      "R/LF/01_LF_ice_thickness_model.R",        "Data/model_outputs/LF_2017_2024_output.csv",
  "GL4",     "R/GL4/01_GL4_ice_model.R",                "Data/model_outputs/GL4_2014_2025_output.csv",
  "LOC",     "R/LOC/01_LOC_ice_model.R",                "Data/model_outputs/LOC_2017_2024_output.csv"
)
# Daily ice temperature profiles, saved by the rerun-models chunk below
lakes$profile_file <- sprintf("Data/model_outputs/%s_thermal_profile.rds", lakes$lake_key)
```

```{r rerun-models}
# Re-run every lake's ice-thickness model so the report always reflects the
# current R/functions/functions.R and CONSTANTS / LAKE_CONFIGS. Each 01_ script
# sources its own 00_ data-prep script, runs the model, and overwrites its CSV
# in Data/model_outputs/. Set `rerun_models: false` in the YAML header (or
# Knit > Knit with Parameters) to skip this and reuse the saved outputs.
rerun_models <- if (exists("params")) isTRUE(params$rerun_models) else TRUE

# The output CSVs can't hold the ice temperature profiles (they are list
# columns), so after each run save one profile per day to an .rds file in
# long format: time, depth (m below the ice surface), temp_C.
save_thermal_profile <- function(res, key) {
  if (is.null(res) || !all(c("depth", "temperature") %in% names(res))) {
    warning("No temperature profiles found for ", key, call. = FALSE)
    return(invisible(NULL))
  }
  daily <- res[!duplicated(as.Date(res$time)), c("time", "depth", "temperature")]
  ok <- purrr::map2_lgl(daily$depth, daily$temperature,
                        ~ length(.x) > 1 && length(.x) == length(.y) && !all(is.na(.x)))
  prof <- daily[ok, ] |>
    tidyr::unnest(c(depth, temperature)) |>
    dplyr::transmute(time,
                     depth  = as.numeric(depth),
                     temp_C = as.numeric(temperature) - 273.15)
  saveRDS(prof, file.path(path, lakes$profile_file[lakes$lake_key == key]))
}

run_lake_model <- function(script, key) {
  old_wd <- setwd(path)                # 01_/00_ scripts use paths relative to the repo root
  on.exit(setwd(old_wd), add = TRUE)
  t0  <- Sys.time()
  env <- new.env(parent = globalenv()) # fresh environment per lake so runs can't leak into each other
  status <- tryCatch({
    source(script, local = env)
    "Re-run"
  }, error = function(e) paste("FAILED:", conditionMessage(e)))
  if (startsWith(status, "FAILED")) {
    warning(script, " ", status, call. = FALSE)
  } else {
    # each 01_ script leaves its model output in `results_<lake_key>`
    res <- mget(paste0("results_", key), envir = env, ifnotfound = list(NULL))[[1]]
    tryCatch(save_thermal_profile(res, key),
             error = function(e) warning("Could not save thermal profile for ", key, ": ",
                                         conditionMessage(e), call. = FALSE))
  }
  tibble::tibble(status  = status,
                 minutes = round(as.numeric(difftime(Sys.time(), t0, units = "mins")), 1))
}

run_log <- if (rerun_models) {
  purrr::map2_dfr(lakes$model_script, lakes$lake_key, run_lake_model)
} else {
  tibble::tibble(status  = rep("Not re-run (saved output)", nrow(lakes)),
                 minutes = NA_real_)
}
run_log <- dplyr::bind_cols(lakes, run_log) |>
  dplyr::mutate(output_written = file.mtime(file.path(path, output_file)))
```

```{r load-data}
# Saved model output for each lake (named list: results$ELB, results$GL4, ...)
results <- lakes$output_file |>
  purrr::map(~ readr::read_csv(file.path(path, .x), show_col_types = FALSE)) |>
  purrr::set_names(lakes$lake_key)

# Matching ice-thickness observations for each lake, clipped to that lake's
# modelled period. load_ice_observations() applies the same lake/location
# filters as the 00_ prep scripts, but needs no met data.
ice_obs <- purrr::imap(results, function(res, key) {
  load_ice_observations(key, project_path = path,
                        time_range = range(res$time, na.rm = TRUE))
})

# Plot one lake: model vs observations. Saved CSVs don't keep the
# temperature-profile list columns, so that panel is switched off.
plot_lake <- function(key) {
  res <- results[[key]]
  plot_ice_model(
    res,
    ice_thickness     = ice_obs[[key]],
    title             = LAKE_CONFIGS[[key]]$lake_name,
    subtitle          = sprintf("%s – %s | %s",
                                format(min(res$time), "%Y-%m-%d"),
                                format(max(res$time), "%Y-%m-%d"),
                                if (nrow(ice_obs[[key]]) > 0)
                                  paste(nrow(ice_obs[[key]]), "ice-thickness observations")
                                else "no ice-thickness observations yet"),
    plot_temp_profile = FALSE
  )
}
```

## Model runs

```{r run-status}
n_failed <- sum(startsWith(run_log$status, "FAILED"))
run_log |>
  dplyr::transmute(Lake = purrr::map_chr(lake_key, ~ LAKE_CONFIGS[[.x]]$lake_name),
                   Script = model_script,
                   Status = status,
                   `Run time (min)` = minutes,
                   `Output last written` = format(output_written, "%Y-%m-%d %H:%M")) |>
  knitr::kable()
```

`r if (n_failed > 0) sprintf("**Warning: %d model run(s) failed. Those lakes are shown using their previously saved output (see the table above).**", n_failed) else ""`

## Ice thickness: all lakes

```{r all-lakes, fig.height = 13}
model_daily <- purrr::imap_dfr(results, function(res, key) {
  res |>
    dplyr::mutate(day = lubridate::floor_date(time, "day")) |>
    dplyr::group_by(day) |>
    dplyr::summarise(thickness = mean(thickness, na.rm = TRUE), .groups = "drop") |>
    dplyr::mutate(lake = LAKE_CONFIGS[[key]]$lake_name)
})
obs_all <- purrr::imap_dfr(ice_obs, ~ dplyr::mutate(.x, lake = LAKE_CONFIGS[[.y]]$lake_name))

lake_levels <- purrr::map_chr(lakes$lake_key, ~ LAKE_CONFIGS[[.x]]$lake_name)
model_daily$lake <- factor(model_daily$lake, levels = lake_levels)
obs_all$lake     <- factor(obs_all$lake,     levels = lake_levels)

ggplot(model_daily, aes(day, thickness)) +
  geom_line(colour = "#2166ac", linewidth = 0.6) +
  geom_point(data = obs_all, aes(date_time, z_water_m),
             colour = "#d6604d", size = 1.4, alpha = 0.8) +
  facet_wrap(~ lake, ncol = 1, scales = "free") +
  labs(x = NULL, y = "Ice thickness (m)",
       caption = "Line = modelled (daily mean); points = observed") +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold", hjust = 0),
        panel.grid.minor = element_blank())
```

## Individual lakes {.tabset}

### West Lake Bonney
```{r wlb}
plot_lake("WLB")
```

### East Lake Bonney
```{r elb}
plot_lake("ELB")
```

### Lake Hoare
```{r lh}
plot_lake("LH")
```

### Lake Fryxell
```{r lf}
plot_lake("LF")
```

### Green Lake 4
```{r gl4}
plot_lake("GL4")
```

### The Loch
```{r loc}
plot_lake("LOC")
```

## Ice thermal profiles

The model solves heat diffusion through the ice column on a `r CONSTANTS$dx` m
grid. The top node is held at the air temperature and the bottom node at the
freezing point (0 °C). These plots use one modelled profile per day, saved
when the models are re-run at the top of this report. Depth is measured down
from the ice surface.

```{r profile-load}
lake_names    <- purrr::map_chr(LAKE_CONFIGS, "lake_name")
profile_paths <- file.path(path, lakes$profile_file)
have_profile  <- file.exists(profile_paths)

profiles <- purrr::map2_dfr(profile_paths[have_profile], lakes$lake_key[have_profile],
                            ~ dplyr::mutate(readRDS(.x), lake_key = .y))

# Profiles saved before the output CSV was last rewritten are out of date
stale <- lakes$lake_key[have_profile][
  file.mtime(profile_paths[have_profile]) <
    file.mtime(file.path(path, lakes$output_file[have_profile])) - 60]

if (nrow(profiles) > 0) {
  dz <- CONSTANTS$dx
  profiles <- profiles |>
    dplyr::mutate(
      lake   = factor(lake_names[lake_key], levels = lake_levels),
      depth  = round(depth / dz) * dz,          # snap to the model grid
      month  = lubridate::month(time),
      region = ifelse(lake_key %in% c("GL4", "LOC"), "Colorado Rockies", "McMurdo Dry Valleys"),
      # season in each lake's own hemisphere (Dry Valleys summer = Dec–Feb)
      nh_month = ifelse(region == "McMurdo Dry Valleys", (month + 5) %% 12 + 1, month),
      season = factor(c("Winter", "Winter", "Spring", "Spring", "Spring", "Summer",
                        "Summer", "Summer", "Autumn", "Autumn", "Autumn", "Winter")[nh_month],
                      levels = c("Winter", "Spring", "Summer", "Autumn"))
    )
}
season_cols <- c(Winter = "#0072B2", Spring = "#009E73", Summer = "#E69F00", Autumn = "#CC79A7")
```

`r if (!all(have_profile)) sprintf("**No saved profiles for %s.** Knit once with <code>rerun_models: true</code> to create them.", paste(lake_names[lakes$lake_key[!have_profile]], collapse = ", ")) else ""`
`r if (length(stale) > 0) sprintf("**The profiles for %s are older than their model output**, so they may not match the rest of this report. Re-knit with <code>rerun_models: true</code> to refresh them.", paste(lake_names[stale], collapse = ", ")) else ""`

### Temperature through the ice column over time

```{r profile-heatmap, fig.height = 15, eval = nrow(profiles) > 0}
# Weekly means keep the plot readable over 7–11 years of daily profiles
prof_week <- profiles |>
  dplyr::mutate(week = as.Date(lubridate::floor_date(time, "week"))) |>
  dplyr::group_by(lake, week, depth) |>
  dplyr::summarise(temp_C = mean(temp_C), .groups = "drop")

ggplot(prof_week, aes(week, depth, fill = temp_C)) +
  geom_tile(width = 7, height = dz) +
  scale_y_reverse(expand = c(0, 0)) +
  scale_x_date(expand = c(0, 0), date_breaks = "1 year", date_labels = "%Y") +
  scale_fill_viridis_c(option = "mako", name = "Ice\ntemperature\n(°C)") +
  facet_wrap(~ lake, ncol = 1, scales = "free") +
  labs(x = NULL, y = "Depth below ice surface (m)",
       title = "Modelled ice temperature with depth through time",
       subtitle = "Weekly mean of daily profiles. The bottom edge of the colour is the ice base; gaps are open water.") +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold", hjust = 0),
        panel.grid = element_blank(),
        legend.position = "right")
```

### Seasonal mean profiles

```{r profile-seasonal, fig.height = 8, eval = nrow(profiles) > 0}
# Mean profile per season, with the 10th–90th percentile of daily values.
# Depths with fewer than 20 daily profiles in a season (the thickest, least
# frequent ice) are dropped so the means aren't based on a handful of days.
prof_season <- profiles |>
  dplyr::group_by(lake, season, depth) |>
  dplyr::summarise(lo = unname(quantile(temp_C, 0.10)),
                   hi = unname(quantile(temp_C, 0.90)),
                   temp_C = mean(temp_C), n = dplyr::n(), .groups = "drop") |>
  dplyr::filter(n >= 20) |>
  dplyr::arrange(lake, season, depth)

ggplot(prof_season, aes(temp_C, depth, colour = season, fill = season)) +
  geom_vline(xintercept = 0, colour = "grey70") +
  geom_ribbon(aes(xmin = lo, xmax = hi), alpha = 0.15, colour = NA, orientation = "y") +
  geom_path(linewidth = 0.9) +
  scale_y_reverse() +
  scale_colour_manual(values = season_cols, name = "Season\n(local hemisphere)") +
  scale_fill_manual(values = season_cols, guide = "none") +
  facet_wrap(~ lake, ncol = 3, scales = "free_y") +
  labs(x = "Ice temperature (°C)", y = "Depth below ice surface (m)",
       title = "Seasonal mean temperature profiles",
       subtitle = "Line = seasonal mean; band = 10th–90th percentile of daily profiles.\nDry Valleys summer = Dec–Feb; Rockies summer = Jun–Aug (mostly open water).") +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold", hjust = 0),
        panel.grid.minor = element_blank(),
        legend.position = "top")
```

```{r profile-table, eval = nrow(profiles) > 0}
profiles |>
  dplyr::group_by(lake, season, time) |>
  dplyr::summarise(t_mean = mean(temp_C), .groups = "drop") |>   # depth-averaged, per day
  dplyr::group_by(Lake = lake, season) |>
  dplyr::summarise(v = sprintf("%.1f", mean(t_mean)), .groups = "drop") |>
  tidyr::pivot_wider(names_from = season, values_from = v, values_fill = "–") |>  # – = no ice that season
  dplyr::left_join(
    profiles |>
      dplyr::group_by(Lake = lake) |>
      dplyr::summarise(`Coldest (°C)` = sprintf("%.1f", min(temp_C)),
                       `Max thickness on grid (m)` = sprintf("%.1f", max(depth))),
    by = "Lake") |>
  knitr::kable(caption = "Depth-averaged ice temperature by season (°C), coldest modelled temperature, and deepest grid node")
```

Because the top node is set to the air temperature, the near-surface part of
these profiles follows the air temperature directly. Once the model solves for
the ice surface temperature, the top of the profile will reflect the surface
energy balance instead.

## Surface energy fluxes

All fluxes are in W m⁻² and are calculated only for **ice-covered** time steps,
so the open-water season at Green Lake 4 and The Loch is left out. Signs follow
the model's surface energy balance:

*net surface flux = absorbed SW + net LW + sensible + latent − Q~c~*

Positive values add energy to the ice surface and can drive melt; negative
values remove energy. The conductive term is shown as it enters the balance
(−Q~c~), so its sign is comparable with the other fluxes.

```{r flux-prep}
flux_vars <- c(SW_abs       = "Absorbed shortwave",
               LW_net       = "Net longwave",
               sensible_Q   = "Sensible heat",
               latent_Q     = "Latent heat",
               conductive_Q = "Conductive (−Qc)")
# Okabe–Ito hues: colour-blind safe; every chart also has a legend and/or a table
flux_cols <- setNames(c("#E69F00", "#0072B2", "#D55E00", "#009E73", "#CC79A7"),
                      flux_vars)

region_of    <- c(WLB = "McMurdo Dry Valleys", ELB = "McMurdo Dry Valleys",
                  LH  = "McMurdo Dry Valleys", LF  = "McMurdo Dry Valleys",
                  GL4 = "Colorado Rockies",    LOC = "Colorado Rockies")
lake_name_of <- purrr::map_chr(LAKE_CONFIGS, "lake_name")

# Long table: one row per lake x time step x flux, ice-covered steps only.
# conductive_Q is negated so it carries the sign it has in the balance.
flux_long <- purrr::imap_dfr(results, function(res, key) {
  if (!"phase" %in% names(res)) res$phase <- "ice"   # perennial lakes: always ice
  res |>
    dplyr::filter(phase == "ice") |>
    dplyr::mutate(conductive_Q = -conductive_Q) |>
    dplyr::select(time, dplyr::all_of(names(flux_vars))) |>
    tidyr::pivot_longer(-time, names_to = "flux_key", values_to = "value") |>
    dplyr::filter(!is.na(value)) |>
    dplyr::mutate(lake_key = key)
}) |>
  dplyr::mutate(
    lake   = factor(lake_name_of[lake_key], levels = lake_levels),
    region = factor(region_of[lake_key],
                    levels = c("McMurdo Dry Valleys", "Colorado Rockies")),
    flux   = factor(flux_vars[flux_key], levels = flux_vars)
  )

# Whole-record statistics per lake and flux
flux_summary <- flux_long |>
  dplyr::group_by(region, lake, lake_key, flux) |>
  dplyr::summarise(mean_signed = mean(value),
                   mean_abs    = mean(abs(value)),
                   p05         = unname(quantile(value, 0.05)),
                   p95         = unname(quantile(value, 0.95)),
                   n_steps     = dplyr::n(),
                   first       = min(time),
                   last        = max(time),
                   .groups = "drop") |>
  dplyr::group_by(lake) |>
  dplyr::mutate(share_abs = mean_abs / sum(mean_abs)) |>
  dplyr::ungroup()

flux_net <- flux_summary |>
  dplyr::group_by(region, lake, lake_key) |>
  dplyr::summarise(net = sum(mean_signed), .groups = "drop")

# Monthly climatology. Months with < 10 days of ice cover across the whole
# record (freeze-up / break-up months at the seasonal lakes) are dropped, and
# months with no ice are kept as NA so lines break rather than bridge the gap.
min_steps_month <- 24 * 10
flux_monthly <- flux_long |>
  dplyr::mutate(month = lubridate::month(time)) |>
  dplyr::group_by(region, lake, lake_key, flux, month) |>
  dplyr::summarise(lo    = unname(quantile(value, 0.10)),
                   hi    = unname(quantile(value, 0.90)),
                   value = mean(value),
                   n     = dplyr::n(), .groups = "drop") |>
  dplyr::filter(n >= min_steps_month) |>
  tidyr::complete(tidyr::nesting(region, lake, lake_key), flux, month = 1:12)

flux_theme <- theme_minimal(base_size = 12) +
  theme(strip.text       = element_text(face = "bold", hjust = 0),
        panel.grid.minor = element_blank(),
        legend.position  = "top")
```

### Flux magnitude across sites

```{r flux-magnitude, fig.height = 6}
ggplot(flux_summary, aes(mean_abs, forcats::fct_rev(lake), colour = flux)) +
  geom_point(size = 3.5, alpha = 0.9) +
  scale_x_log10(breaks = 10^(-3:2),
                labels = c("0.001", "0.01", "0.1", "1", "10", "100")) +
  scale_colour_manual(values = flux_cols, name = NULL) +
  facet_grid(region ~ ., scales = "free_y", space = "free_y") +
  labs(x = "Mean absolute flux (W m⁻², log scale)", y = NULL,
       title = "How large is each flux?",
       subtitle = "Mean |flux| over all ice-covered hours. Each gridline is a factor of 10.") +
  flux_theme +
  guides(colour = guide_legend(nrow = 1))
```

```{r flux-magnitude-table}
flux_summary |>
  dplyr::mutate(cell = sprintf("%.3g (%.0f%%)", mean_abs, 100 * share_abs)) |>
  dplyr::select(Region = region, Lake = lake, flux, cell) |>
  tidyr::pivot_wider(names_from = flux, values_from = cell) |>
  knitr::kable(caption = "Mean |flux| in W m⁻² (share of the lake's total flux magnitude)",
               align = "llrrrrr")
```

### Energy sources and sinks

```{r flux-signed, fig.height = 6}
ggplot(flux_summary, aes(mean_signed, forcats::fct_rev(lake), fill = flux)) +
  geom_col(width = 0.65, colour = "white", linewidth = 0.4) +
  geom_vline(xintercept = 0, colour = "grey30") +
  geom_point(data = flux_net, aes(net, forcats::fct_rev(lake)),
             inherit.aes = FALSE, shape = 23, size = 3.5,
             fill = "black", colour = "white") +
  scale_fill_manual(values = flux_cols, name = NULL) +
  facet_grid(region ~ ., scales = "free_y", space = "free_y") +
  labs(x = "Mean flux (W m⁻²)", y = NULL,
       title = "What adds and removes energy at the ice surface?",
       subtitle = "Bars to the right add energy, bars to the left remove it. Diamond = mean net surface flux.") +
  flux_theme +
  guides(fill = guide_legend(nrow = 1))
```

```{r flux-signed-table}
flux_summary |>
  dplyr::mutate(cell = sprintf("%+.2f", mean_signed)) |>
  dplyr::select(Region = region, Lake = lake, flux, cell) |>
  tidyr::pivot_wider(names_from = flux, values_from = cell) |>
  dplyr::left_join(dplyr::transmute(flux_net, Lake = lake,
                                    `Net surface flux` = sprintf("%+.2f", net)),
                   by = "Lake") |>
  knitr::kable(caption = "Mean signed flux in W m⁻² (ice-covered hours)",
               align = "llrrrrrr")
```

### Seasonal cycle

```{r flux-seasonal, fig.height = 10}
ggplot(flux_monthly, aes(month, value, colour = flux)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_line(linewidth = 0.8, na.rm = TRUE) +
  geom_point(size = 1.6, na.rm = TRUE) +
  scale_x_continuous(breaks = 1:12, labels = substr(month.abb, 1, 1)) +
  scale_colour_manual(values = flux_cols, name = NULL) +
  facet_wrap(~ lake, ncol = 2) +
  labs(x = NULL, y = "Monthly mean flux (W m⁻²)",
       title = "Monthly flux climatology",
       subtitle = "Dry Valleys summer = Dec–Jan; Rockies summer = Jun–Jul.\nGaps = open water (< 10 days of ice in that month).") +
  flux_theme +
  guides(colour = guide_legend(nrow = 1))
```

### Comparing the regions

```{r region-numbers}
reg <- flux_summary |>
  dplyr::group_by(region, flux) |>
  dplyr::summarise(mean_abs = mean(mean_abs), mean_signed = mean(mean_signed),
                   .groups = "drop")
rv <- function(r, f, what = "mean_abs") {
  reg[[what]][reg$region == r & reg$flux == flux_vars[[f]]]
}
mdv <- "McMurdo Dry Valleys"; rmt <- "Colorado Rockies"
```

- **Radiation and conduction dominate everywhere.** At every lake, absorbed
  shortwave, net longwave and conduction together make up more than
  `r sprintf("%.0f%%", 100 * min(dplyr::summarise(dplyr::group_by(dplyr::filter(flux_summary, flux %in% flux_vars[c("SW_abs","LW_net","conductive_Q")]), lake), s = sum(share_abs))$s))`
  of the total flux magnitude. Sensible heat is about 0.5 W m⁻² and latent
  heat about 0.001–0.01 W m⁻², which is one to four orders of magnitude
  smaller.
- **The Dry Valley lakes run on stronger radiation.** Mean absorbed SW is
  `r sprintf("%.0f", rv(mdv, "SW_abs"))` W m⁻² in the Dry Valleys versus
  `r sprintf("%.0f", rv(rmt, "SW_abs"))` W m⁻² in the Rockies, and net LW is
  `r sprintf("%.0f", rv(mdv, "LW_net", "mean_signed"))` versus
  `r sprintf("%.0f", rv(rmt, "LW_net", "mean_signed"))` W m⁻². At the
  Antarctic lakes net LW is a sink in every month. It is largest in summer,
  when it partly offsets the 24-hour sun. Absorbed SW falls to zero through
  the polar night (May–Aug).
- **The Rockies lakes are frozen only during the low-sun half of the year.**
  Their ice season runs from about Oct/Nov to Jun/Jul, so they skip the
  months when SW would be largest. Net LW at these lakes switches from a
  winter sink to a spring source as the air warms, and that spring LW gain
  together with SW drives ice-out.
- **Conduction moves energy around but adds almost none.** Averaged over the
  record, −Q~c~ is near zero at every lake. Its hourly swings, however, are
  as large as the radiative terms (5th–95th percentile roughly ±25–45 W m⁻²):
  conduction carries heat into and out of the ice through the daily cycle
  but adds or removes very little over a full season.

### Flux budget by lake {.tabset}

```{r flux-describe-fn}
fmt <- function(x, d = 1) formatC(x, format = "f", digits = d, big.mark = ",")
fmt_flux <- function(x) ifelse(abs(x) < 0.1, formatC(x, format = "g", digits = 2),
                               fmt(x))

describe_fluxes <- function(key) {
  s   <- dplyr::filter(flux_summary, lake_key == key)
  m   <- dplyr::filter(flux_monthly, lake_key == key, !is.na(value))
  net <- flux_net$net[flux_net$lake_key == key]
  get <- function(f, col) s[[col]][s$flux == flux_vars[[f]]]

  ranked  <- dplyr::arrange(s, dplyr::desc(mean_abs))
  rank_tx <- paste0(ranked$flux[1:3], " (", fmt_flux(ranked$mean_abs[1:3]),
                    " W m⁻², ", sprintf("%.0f%%", 100 * ranked$share_abs[1:3]), ")")
  src  <- dplyr::slice_max(s, mean_signed, n = 1)
  sink <- dplyr::slice_min(s, mean_signed, n = 1)

  sw_m <- dplyr::filter(m, flux == flux_vars[["SW_abs"]])
  lw_m <- dplyr::filter(m, flux == flux_vars[["LW_net"]])
  net_m <- m |> dplyr::group_by(month) |> dplyr::summarise(net = sum(value), .groups = "drop")
  pos_months <- month.abb[net_m$month[net_m$net > 0]]
  ice_months <- month.abb[sort(unique(m$month))]

  n_days <- s$n_steps[1] / 24
  paste0(
    "**", lake_name_of[[key]], "** (", region_of[[key]], "). This summary covers ",
    fmt(n_days, 0), " ice-covered days between ", format(s$first[1], "%b %Y"),
    " and ", format(s$last[1], "%b %Y"),
    if (length(ice_months) < 12)
      paste0(" (ice in ", paste(ice_months, collapse = ", "), ")") else "",
    ".\n\n",
    "- **Largest fluxes by magnitude:** ", paste(rank_tx, collapse = "; "), ".\n",
    "- **Main energy source:** ", src$flux, " (mean ", sprintf("%+.1f", src$mean_signed),
    " W m⁻²). **Main sink:** ", sink$flux, " (", sprintf("%+.1f", sink$mean_signed),
    " W m⁻²). The mean net surface flux is ", sprintf("%+.1f", net), " W m⁻²",
    if (net < 0) ", so on average the surface loses energy and surface melt is limited to the warmest hours."
    else ", so on average the surface gains energy, which favours surface melt.", "\n",
    "- **Seasonality:** absorbed SW peaks in ",
    month.abb[sw_m$month[which.max(sw_m$value)]], " (", fmt(max(sw_m$value)), " W m⁻²)",
    " and net LW is most negative in ", month.abb[lw_m$month[which.min(lw_m$value)]],
    " (", fmt(min(lw_m$value)), " W m⁻²). ",
    if (length(pos_months) > 0)
      paste0("The monthly mean net flux is positive (melt-favouring) in ",
             paste(pos_months, collapse = ", "), ".")
    else "The monthly mean net flux is negative in every month.", "\n",
    "- **Conduction:** the mean is ", sprintf("%+.2f", get("conductive_Q", "mean_signed")),
    " W m⁻², but hourly values span ", fmt(get("conductive_Q", "p05")), " to ",
    fmt(get("conductive_Q", "p95")), " W m⁻² (5th–95th percentile).\n",
    "- **Turbulent fluxes:** sensible ", fmt_flux(get("sensible_Q", "mean_signed")),
    " W m⁻² and latent ", fmt_flux(get("latent_Q", "mean_signed")),
    " W m⁻² on average. Together they are ",
    sprintf("%.1f%%", 100 * (get("sensible_Q", "share_abs") + get("latent_Q", "share_abs"))),
    " of the total flux magnitude.\n"
  )
}

# Per-lake monthly panel: mean line + 10–90th percentile band of hourly values
plot_lake_fluxes <- function(key) {
  d <- dplyr::filter(flux_monthly, lake_key == key)
  ggplot(d, aes(month, value, colour = flux, fill = flux)) +
    geom_hline(yintercept = 0, colour = "grey60") +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, colour = NA, na.rm = TRUE) +
    geom_line(linewidth = 0.8, na.rm = TRUE) +
    geom_point(size = 1.4, na.rm = TRUE) +
    scale_x_continuous(breaks = 1:12, labels = substr(month.abb, 1, 1)) +
    scale_colour_manual(values = flux_cols, guide = "none") +
    scale_fill_manual(values = flux_cols, guide = "none") +
    facet_wrap(~ flux, nrow = 1, scales = "free_y") +
    labs(x = NULL, y = "W m⁻²",
         subtitle = "Monthly mean (line) and 10th–90th percentile of hourly values (band). Each panel has its own y-axis.") +
    flux_theme
}

flux_tab <- function(key) {
  cat("\n\n#### ", lake_name_of[[key]], "\n\n", sep = "")
  cat(describe_fluxes(key), "\n\n")
  print(plot_lake_fluxes(key))
  cat("\n\n")
}
```

```{r flux-tabs, results = "asis", fig.height = 3.2, fig.width = 12}
for (key in lakes$lake_key) flux_tab(key)
```
